---
title: "Assessing the single-cell specific abundance of lin-41 and set-3 from early embryonic cells"
author: "Erin Nishimura"
date: "February 9, 2026"
output: 
  github_document:
    toc: true
    toc_depth: 4
---


The purpose of this script is to assess the abundance of lin-41 mRNA and set-3 mRNA, not on a whole embryo level, but on a cell-specific level. The wormLib package currently tabulates smFISH microscopy images for abundance based on individual blastomeres for 2-cell and 4-cell stage embryos. Beyond this, the cells are too difficult to differentiate without additional markers.

This project is being conducted in the revisions of the Spike et al manuscript.

This analysis will appear in Figure 2 of Spike et al.

# Load required packages

  * tidyverse
  * gridExtra
  

```{r echo=FALSE, warning=FALSE, message=FALSE} 

library(tidyverse)
library(gridExtra)

```


# Import smFISH abundance data

  * This is output from wormLib (bigFISH)

```{r importFISH, echo=TRUE, warning = FALSE, message = FALSE}

# read in the mRNA abundance data from smFISH quantifications:
FISH <- read.csv(file = "../01_input/combined_cell_data.csv")

# peek at the data:
head(FISH)

# data structure:
str(FISH)

```

# Reformat the smFISH abundance data

  1. There is an issue where ABp cells are listed as ABb cells. Fix it.
  2. Pivot longer to unify mRNA columns
  3. Add lineage info 
  4. Select only desired columns 
  5. Set factors 

```{r, echo=TRUE, warning = FALSE, message = FALSE}

# === 1. There is an issue where ABp cells are listed as ABb cells. Fix it. ===
FISH <- FISH %>%
  mutate(label = case_when(label == "ABb" ~ "ABp" , .default = label))
unique(FISH$label)

# === 2. Pivot longer to unify mRNA columns ===
FISH_long <- FISH %>%
  pivot_longer(
    cols = ends_with("_mRNA"),
    names_to = "mRNA_name",
    values_to = "abundance"
  )

# === 3. Add lineage info ===
FISH_long <- FISH_long %>%
  mutate(
    lineage = case_when(
      label %in% c("AB", "ABa", "ABp", "EMS") ~ "Somatic",
      label %in% c("P1", "P2", "P3", "P4") ~ "Germline",
      TRUE ~ NA_character_
    ),
    lineage = factor(lineage, levels = c("Somatic", "Germline"))  # Force order
  )

# === 4. Select only desired columns ===
FISH_filtered <- FISH_long %>%
  select(ImageID, label, rep, cell_stage, mRNA_name, abundance, lineage) %>%
  filter(!is.na(abundance))

# === 5. Set factors ===
FISH_filtered$ImageID <- factor(FISH_filtered$ImageID)
FISH_filtered$label <- factor(FISH_filtered$label, level = c("AB", "P1", "ABa", "ABp", "EMS", "P2"))
FISH_filtered$mRNA_name <- factor(FISH_filtered$mRNA_name)
FISH_filtered$cell_stage <- factor(FISH_filtered$cell_stage)

# Final dataset
head(FISH_filtered)
str(FISH_filtered)

```

# Import the cell size metrics

  * These were part of the WormLib (bigFISH) output
  * Import cell sizes
  * Calculate mean cell volume per cell type
  * Merge cell volume data with the FISH Abundance data

```{r importsizes, echo=TRUE, warning = FALSE, message = FALSE}

# Read in cell size estimates
cellsizes <- read.table(file = "../01_input/260209_merged_output_cellsizes_shorter.txt", header = TRUE)
head(cellsizes)

# Merge the FISH Abundance with the cell sizes data:
FISHcounts_with_sizes <- left_join(FISH_filtered, cellsizes)

#print(FISHcounts_with_sizes, n = 214, keep = TRUE)

# Calculate the mean size of each cell type
size_per_cell <- FISHcounts_with_sizes %>%
  group_by(label) %>%
  summarise(circumference = mean(perimeter, na.rm = TRUE)) %>%
  mutate(volume = circumference^3 / 6*(pi^2) )

# append the mean cell circumferences and volumes associated with each cell type back into the FISH_counts_with_sizes data:
FISHcounts_total_sizes <- left_join(FISHcounts_with_sizes, size_per_cell)
FISHcounts_total_sizes$ImageID <- factor(FISHcounts_total_sizes$ImageID)
FISHcounts_total_sizes$label <- factor(FISHcounts_total_sizes$label, level = c("AB", "P1", "ABa", "ABp", "EMS", "P2"))

head(FISHcounts_total_sizes)
dim(FISHcounts_total_sizes)


```

# Filter by lin-41 counts and plot

```{r lin41, echo=TRUE, warning = FALSE, message = FALSE}

# Filter by lin-41
lin41_selection <- FISHcounts_total_sizes %>%
  filter(mRNA_name == "lin41_mRNA")

# Are there any unmatched entries?
unmatched <- lin41_selection %>%
  filter(is.na(perimeter))
length(unique(unmatched$ImageID))

# 0

# Divide the abundance by the volume:
scaled_lin41_selection <- lin41_selection %>%
  mutate(abundance_by_vol = abundance / volume)

print(scaled_lin41_selection, width = Inf)

# Plot abundance scaled by estimated cell type volume 

u <- ggplot(scaled_lin41_selection, aes(x=label, y=abundance_by_vol, fill=lineage)) + 
    geom_boxplot() +
    geom_jitter(shape=16, position=position_jitter(0.2), size = 1) +
    labs(title = "lin-41 mRNA abundance by volume",
       x = "Cell Identity",
       y = "lin-41 mRNA, relative concentration per volume")
u

```


# Filter by set-3 counts and plot

```{r set3, echo=TRUE, warning = FALSE, message = FALSE}

# Filter by lin-41
set3_selection <- FISHcounts_total_sizes %>%
  filter(mRNA_name == "set3_mRNA")

#dim(set3_selection)
# 106 rows

# Divide the abundance by the volume:
scaled_set3_selection <- set3_selection %>%
  mutate(abundance_by_vol = abundance / volume)

#print(scaled_set3_selection, width = Inf)

# Plot abundance scaled by estimated cell type volume 

v <- ggplot(scaled_set3_selection, aes(x=label, y=abundance_by_vol, fill=lineage)) + 
    geom_boxplot() +
    geom_jitter(shape=16, position=position_jitter(0.2), size = 1) +
    labs(title = "set-3 mRNA abundance by volume",
       x = "Cell Identity",
       y = "set-3 mRNA, relative concentration per volume")
v

```

# Print and save relevant plots

This will be included as Figure 2C

```{r printplots, , echo = TRUE, message = FALSE, warning = FALSE}

# Set today's date
today <- format(Sys.Date(), "%y%m%d")

# set plotname
filename <- paste("../03_output/", today, "_scaled_boxplots.pdf", sep = "")

# save pdf
pdf(filename, width = 10, height = 6)

grid.arrange(u, v, ncol = 2)

dev.off()


```

# Export data tables

```{r printplots, , echo = TRUE, message = FALSE, warning = FALSE}

# Merge teh lin-41 and set-3 data together
lin41_and_set3_selection <- rbind(scaled_lin41_selection, scaled_set3_selection)

lin41_selection$mRNA_name
dim(lin41_and_set3_selection)
colnames(lin41_and_set3_selection)
head(lin41_and_set3_selection)
# Split out the date:
lin41_and_set3_selection$rep

lin41_and_set3_selection2 <- lin41_and_set3_selection %>%
  mutate(date = substr(lin41_and_set3_selection$ImageID, 0, 6)) %>%
  select(ImageID, date, mRNA_name, cell_stage, label, lineage, abundance, perimeter, major_axis_length, minor_axis_length, circumference, volume, abundance_by_vol)

lin41_and_set3_selection2$date <- as.factor(lin41_and_set3_selection2$date)
lin41_and_set3_selection2$mRNA_name <- droplevels(lin41_and_set3_selection2$mRNA_name )

lin41_and_set3_selection2

# Calculate the n-values per stage and transcript type
nvalues <- lin41_and_set3_selection2 %>%
  select(cell_stage, mRNA_name) 

# Convert to a data frame
ntable <- table(nvalues)
as.data.frame(ntable)

# Calculate the rep-values per stage and transcript type
repvalues <- lin41_and_set3_selection2 %>%
  select(date, mRNA_name)

# Convert to a data frame
reptable <- table(repvalues)
as.data.frame(reptable)


# Set today's date
today <- format(Sys.Date(), "%y%m%d")

# set raw data filename - lin-41 and set-3
filename <- paste("../04_data/", today, "_dataTables_Figure2C.txt", sep = "")
# save raw data - lin-41 and set-3
write.table(lin41_and_set3_selection2, file = filename, quote = FALSE)

# set filename - n-values
filename2 <- paste("../04_data/", today, "_n_values_Figure2C.txt", sep = "")
# save n-values
write.table(as.data.frame(ntable), file = filename2, quote = FALSE)

# set filename - n-values
filename3 <- paste("../04_data/", today, "_rep_values_Figure2C.txt", sep = "")
# save n-values
write.table(as.data.frame(reptable), file = filename3, quote = FALSE)



```


# Session info

```{r}

sessionInfo()

```

