---
title: "Analyze and plot total RNA counts in spn-4 wild-type v. spn-4 null strains"
output: 
  github_document:
    toc: true
    toc_depth: 4
---

# Load Required packages

  - tidyverse
  - dplyr
  - ggplot2
  - ggthemes
  - ggpubr
  - rstatix
  - pheatmap

```{r setup,  echo = FALSE, warning = FALSE, message = FALSE}
print("script initiated on", quote = FALSE)
Sys.time()
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(rstatix)
library(pheatmap)


```

# Import the RNA molecule counts from smFISH/smiFISH images

Import the data. The file is 'all_spn4_smiFISH_quantification_250902.csv' and it is included as a Supplemental Data table.

These quantifications generated from smFISH and smiFISH microscopy images performed on strain DG5256, a spn-4 balancer strain and on N2, the standard strain. In this DG5256, embryos that were produced by spn-4::gfp mothers fluoresce green and are labeled here as spn4_plus. In DG5256, embryos that were produced from spn-4 null mothers do not fluoresce green and are labeled here as spn4_minus. Embryos from the strain N2 are labeled spn4_wt. 

```{r import, echo = TRUE, warning = FALSE}
# import the data files
mywd <- "~/Library/CloudStorage/Dropbox/github/SPN4_maternal_mRNA/08_spn4_null"
setwd(mywd)
all_spn4_dataset <- read.table(file = "01_input/all_spn4_smiFISH_quantification_250902.csv", header = TRUE, sep = ",")

#Set the cell stage and genotypes as ordered factors
all_spn4_dataset$cell_stage <- factor(all_spn4_dataset$cell_stage, levels = c("2-cell", "4-cell", "8-11-cell", "12-15-cell", "16-28-cell"))
all_spn4_dataset$spn4 <- factor(all_spn4_dataset$spn4, levels = c("spn4_plus", "spn4_minus", "spn4_wt"))

#EDA on the spn4 dataset:
dim(all_spn4_dataset)
head(all_spn4_dataset)

```


# Data munging

Group the data by transcript and cell type and calculate the total number of RNA molecules (and standard deviation) for each of those condition combinations. 

```{r munging, echo = TRUE, warning = FALSE, message = FALSE}

# Make quant table: 1) group by genotype & cell stage and 2) calculate mean & sd, optional 
allmRNAs_quantTable <- all_spn4_dataset %>%
  dplyr::group_by(transcript, spn4, cell_stage) %>%
  dplyr::summarise(mean = mean(total_no_RNA), sd = sd(total_no_RNA))

#head(allmRNAs_quantTable)
#str(allmRNAs_quantTable)

```


# Calculate Adjusted P-values using T Tests & P-Adjust

T-tests were used to determine significant differences. I am primarily interested in contrasting the SPN-4 minus condition versus the SPN-4 wild type to determine how SPN-4 influences mRNA abundance. In addition, I want to focus in on how SPN-4 plus contrasts from wild-type conditions to determine how the SPN-4::GFP fusion construct acts as a hypomorph compared to the wild type SPN-4. 

T-tests were performed with setting var.equal = FALSE (default) such that Welch approximation will be used to take into account variances of unequal values.

I tested both Bonferonni correction and Benjamini-Hochberg multiple test corrections. The Bonferroni correction was too severe, but B&H worked well. I will go with the B&H version here.

```{r ttests, echo = TRUE, warning = FALSE}

# Use the following dataframe:
# all_spn4_dataset

# Calculate t-test significance. 
# Adjust p-values by BH correction
stat.test_BH <- all_spn4_dataset %>%
  group_by(transcript, cell_stage) %>%
  t_test(total_no_RNA ~ spn4) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  add_xy_position(x = "cell_stage" , fun = "max", dodge = 0.8 , scales = "free")

# Set some y-values that will be used to plot the adjusted p-values on the plot. In each case, 
# spn-4 minus v. spn-4 plus will be on TOP, spn-4 minus versus N2 will be in the MIDDLE and
# spn4-plus v. N2 will be on the BOTTOM.
stat.test_BH_plus_1000 <- stat.test_BH%>%
  mutate(compare = case_when(group1=="spn4_plus" & group2=="spn4_minus" ~  1000,
                           group1=="spn4_minus" & group2=="spn4_wt" ~ 500,
                           group1=="spn4_plus" & group2=="spn4_wt" ~ 0))

stat.test_BH_plus_10000 <- stat.test_BH %>%
  mutate(compare = case_when(group1=="spn4_plus" & group2=="spn4_minus" ~  10000,
                           group1=="spn4_minus" & group2=="spn4_wt" ~ 5000,
                           group1=="spn4_plus" & group2=="spn4_wt" ~ 0))

head(stat.test_BH_plus_1000)

```


## Figure S5: Plot all lineplots - all three genotypes + p-values

Print the lineplots of all the the smFISH/smiFISH data. BH-adjusted, Welch approximated t-test adj-p-values will also be printed on top of the lineplots. Because it was difficult to color these stats, I will print a very different versions.

These will be used in Supplemental Figure 5

In each case...

  - spn-4 minus v. spn-4 plus will be on TOP, 
  - spn-4 minus versus N2 will be in the MIDDLE and
  - spn4-plus v. N2 will be on the BOTTOM.

```{r savePlots1, echo = FALSE, warning = FALSE, message = FALSE}

plotx <- ggplot(allmRNAs_quantTable, aes(x=cell_stage, y=mean, group=spn4, color=spn4, shape = spn4)) + 
  geom_ribbon(aes(ymin=(mean-sd), ymax=(mean+sd), fill = spn4), alpha=0.1, linetype = 0) +
  facet_wrap(.~transcript,scales='free') +
  labs(x = "Cell Stage", y = "Total mRNA molecules") +
  geom_line() + geom_point() +
  theme_linedraw() +
  scale_fill_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
    scale_color_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
  stat_pvalue_manual(stat.test_BH_plus_1000, x = "cell_stage", y.position = "compare", hide.ns = FALSE, label = "p.adj.signif", remove.bracket = TRUE) +
  theme(strip.text = element_text(face = "italic"), 
        panel.grid.major = element_line(colour = "grey"), 
        axis.text.x = element_text(angle=45, hjust = 1))

plotx

# Save plot BH and stats (0 - 1000)
setwd(mywd)
today <- format(Sys.Date(), "%y%m%d")
filename2 <- paste("03_plots/", today, "_smiFISH_lineplots_3strains_BH_1000.pdf", sep = "")
pdf(filename2, width = 10, height = 12)

plotx

dev.off()

# BH and stats (0 - 10000)
setwd(mywd)
getwd()
today <- format(Sys.Date(), "%y%m%d")
filename2 <- paste("03_plots/", today, "_smiFISH_lineplots_3strains_BH_10000.pdf", sep = "")
pdf(filename2, width = 10, height = 12)

ggplot(allmRNAs_quantTable, aes(x=cell_stage, y=mean, group=spn4, color=spn4, shape = spn4)) + 
  geom_ribbon(aes(ymin=(mean-sd), ymax=(mean+sd), fill = spn4), alpha=0.1, linetype = 0) +
  facet_wrap(.~transcript,scales='free') +
  labs(x = "Cell Stage", y = "Total mRNA molecules") +
  geom_line() + geom_point() +
  theme_linedraw() +
  scale_fill_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
  scale_color_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
  stat_pvalue_manual(stat.test_BH_plus_10000, x = "cell_stage", y.position = "compare", hide.ns = FALSE, label = "p.adj.signif", remove.bracket = TRUE) +
  theme(strip.text = element_text(face = "italic"), 
        panel.grid.major = element_line(colour = "grey"), 
        axis.text.x = element_text(angle=45, hjust = 1))

dev.off()

# BH and No Stats
setwd(mywd)
getwd()
today <- format(Sys.Date(), "%y%m%d")
filename2 <- paste("03_plots/", today, "_smiFISH_lineplots_3strains_BH_noStats.pdf", sep = "")
pdf(filename2, width = 10, height = 12)

ggplot(allmRNAs_quantTable, aes(x=cell_stage, y=mean, group=spn4, color=spn4, shape = spn4)) + 
  geom_ribbon(aes(ymin=(mean-sd), ymax=(mean+sd), fill = spn4), alpha=0.1, linetype = 0) +
  facet_wrap(.~transcript,scales='free') +
  labs(x = "Cell Stage", y = "Total mRNA molecules") +
  geom_line() + geom_point() +
  theme_linedraw() +
  scale_fill_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
    scale_color_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_plus" = "#56B4E9", 
                               "spn4_wt" = "#000000"))+
  theme(strip.text = element_text(face = "italic"), 
        panel.grid.major = element_line(colour = "grey"), 
        axis.text.x = element_text(angle=45, hjust = 1))

dev.off()

```

## Figure 3: Plot lineplots of select transcripts with just two genotypes

This plot will be for Figure 3. It will just focus on few of the transcripts and will restrict the lineplots to spn-4 minus and spn-4 wild type.

```{r savePlots2, echo = TRUE, warning = FALSE, message = FALSE}

# Subset the transcripts - lin-41, chs-1, cpg-2, oma-2, mcm-2, and nasp-2
allmRNAs_quantTable_fig3 <- allmRNAs_quantTable %>%
  filter(transcript == "lin-41" | transcript == "chs-1" | transcript == "cpg-2" |  transcript == "oma-2" | transcript == "mcm-2" | transcript == "nasp-2" ) %>%
  filter(spn4 == "spn4_minus" | spn4 == "spn4_wt" )

# Set the order for the transcripts
unique(allmRNAs_quantTable_fig3$transcript)
allmRNAs_quantTable_fig3$transcript <- factor(allmRNAs_quantTable_fig3$transcript, levels = c("lin-41", "chs-1", "cpg-2", "oma-2", "mcm-2", "nasp-2"))

# Filter the statistics
stat_test_BH_fig3 <- stat.test_BH %>%
  filter(transcript == "lin-41" | transcript == "chs-1" | transcript == "cpg-2" |  transcript == "oma-2" | transcript == "mcm-2" | transcript == "nasp-2" )

# Add where the p-adj statistics will be placed on the plot 
stat.test_BH_plus_5 <- stat_test_BH_fig3 %>%
  mutate(compare = case_when(group1=="spn4_plus" & group2=="spn4_minus" ~  5,
                           group1=="spn4_minus" & group2=="spn4_wt" ~ 2.5,
                           group1=="spn4_plus" & group2=="spn4_wt" ~ 0.1))

ploty <- ggplot(allmRNAs_quantTable_fig3, aes(x=cell_stage, y=mean/1000, group=spn4, color=spn4, shape = spn4)) + 
  geom_ribbon(aes(ymin=(mean-sd)/1000, ymax=(mean+sd)/1000, fill = spn4), alpha=0.1, linetype = 0) +
  facet_wrap(.~transcript,scales='free') +
  labs(x = "Cell Stage", y = "Total mRNA molecules") +
  geom_line() + geom_point() +
  theme_linedraw() +
  ylim(0, 28) +
  scale_fill_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_wt" = "#000000"))+
    scale_color_manual(values = c("spn4_minus" = "#E69F00", 
                               "spn4_wt" = "#000000"))+
  stat_pvalue_manual(stat.test_BH_plus_5, x = "cell_stage", y.position = "compare", hide.ns = FALSE, label = "p.adj.signif", remove.bracket = TRUE) +
  theme(strip.text = element_text(face = "italic"), 
        panel.grid.major = element_line(colour = "grey"), 
        axis.text.x = element_text(angle=45, hjust = 1))

ploty

# Plot for Figure 3  - 
# BH and Stats
setwd(mywd)
today <- format(Sys.Date(), "%y%m%d")
filename2 <- paste("03_plots/", today, "_smiFISH_lineplots_2-strains_BH_stats.pdf", sep = "")
pdf(filename2, width = 6, height = 6)

ploty

dev.off()

```

# Figure F3E: Calculate and plot the 4-cell difference values

I tested what would be the best metric of spn-4's impact on mRNA abundance. I tested

  * log2 fold change differences at the 4-cell stage between SPN-4 minus / SPN-4 wt
  * log2 fold change differences at teh 8 to 11-cell stages between SPN-4 minus / SPN-4 wt
  * Area Under the Curve of the difference curves between SPN-4 minus v. SPN-4 wt
  
They all looked very similar. We settled on the 4-cell stage as this was the most relevant to our early stage interest 

```{r 4cell_test, echo = TRUE, warning = FALSE, message = FALSE} 

# Calculate the log2 fold change difference between spn-4 minus and spn-4 wild type at the 4-cell stage
Fourcell_foldChange <- allmRNAs_quantTable %>%
  filter(cell_stage == "4-cell") %>%
  select(transcript, cell_stage, spn4, mean) %>%
  group_by(transcript) %>%
  pivot_wider(names_from = spn4, values_from = mean)%>%
  mutate(spn4_difference_4cell = log(spn4_minus/spn4_wt, base = 2)) %>%
  arrange(desc(spn4_difference_4cell))

# Get data from 250226 Plotting Tatsuya Data
setwd(mywd)
filtered_SPN4 <- read.table(file = "01_input/20250829_smFISH_assayed_transcripts.txt", header = TRUE)
colnames(filtered_SPN4)[2] <- "transcript"

# Merge assayed Genes data frame (RIP-seq) and fold change data frame (smFISH)
joined_filtered <- left_join(Fourcell_foldChange, filtered_SPN4)

# Order by descending 
joined_filtered <- joined_filtered %>%
  arrange(desc(SPN4), desc(spn4_enrichment))

# Save transcript as a rowname
joined_filtered <- as.data.frame(joined_filtered)
rownames(joined_filtered) <- joined_filtered$transcript

## Subset for the RBP states
#joined_filtered[,11:13]
#joined_filtered[,20]
joined_filtered_rpm <- joined_filtered[,c(11:13,20)]

# Create an expression mean data frame (for annotations)
joined_filtered_df_mean <- as.data.frame(joined_filtered[,20])
rownames(joined_filtered_df_mean) <- joined_filtered$transcript

# Plot a heatmap of the IP-seq enrichment values for all three RBPs
v <- pheatmap(joined_filtered_rpm[,c(3,2,1)], 
              scale="none", 
              color = colorRampPalette(c("blue4", "beige", "maroon2"), space = "Lab")(1000),
              cluster_rows=FALSE, 
              cluster_cols=FALSE, 
              border_color = "white", 
              show_rownames = TRUE,
              annotation_row = joined_filtered_df_mean
)

v

# Save plots
# Save the heatmap plot illustrating SPN-4 enrichment:
setwd(mywd)
getwd()
today <- format(Sys.Date(),"%Y%m%d")
file6 <- paste("03_plots/", today, "_heatmap_smFISH_selecteGenes_orderedBySPN4.pdf", sep = "")

pdf(file6, height = 9, width = 6)
v
dev.off()


# Heatmap plot of expression level
w <- pheatmap(joined_filtered$mean, 
              scale="none", 
              color = colorRampPalette(c("darkgreen", "white", "orange"), space = "Lab")(100),
              cluster_rows=FALSE, 
              cluster_cols=FALSE, 
              border_color = "white",
              show_rownames=TRUE)

w

# Save the plot 
today <- format(Sys.Date(),"%Y%m%d")
file7 <- paste("03_plots/", today, "_heatmap_expression_ordered_bySPN4_cat.pdf", sep = "")

pdf(file7, height = 9, width = 6)
w
dev.off()

# Save Barplot of 4-cell stage fold change...
barplot(joined_filtered$spn4_difference_4cell, names.arg = joined_filtered$transcript, 
        las = 2, 
        xlab = "transcript", 
        ylab = "Fold change at 4-cell")

today <- format(Sys.Date(),"%Y%m%d")
today
file8 <- paste("03_plots/", today, "_4-cell_difference_ordered_bySPN4_cat.pdf", sep = "")
pdf(file8, height = 9, width = 6)
barplot(joined_filtered$spn4_difference_4cell, names.arg = joined_filtered$transcript, 
        las = 2, 
        xlab = "transcript", 
        ylab = "Fold change at 4-cell")
dev.off()

```

# Export supplemental data

```{r export_supp, echo = TRUE, warning = FALSE}

setwd(mywd)
file10 <- paste("04_supplemental_data/", today, "_adj_pvalues_w_and_wo_spn4.txt", sep = "")
write.table(as.data.frame(stat.test_BH_plus_1000[ ,c(1:12)]), file = file10, sep = "\t", row.names = FALSE, quote = FALSE)

file11 <- paste("04_supplemental_data/", today, "_means_and_sd_totalRNAs_w_and_wo_spn4.txt", sep = "")
write.table(allmRNAs_quantTable, file = file11, sep = "\t", row.names = FALSE, quote = FALSE)

```

# Capture session and version info

```{r session, echo = TRUE}

setwd(mywd)
file12 <- paste("04_supplemental_data/", today, "_sessionInfo.txt", sep = "")
writeLines(capture.output(sessionInfo()), file12)
```
