---
title: "250717_Parsing_scRNA-seq_Tintori"
output: github_document
---

# 

## Load Packages

The following packages are required:

  * corrplot
  * RColorBrewer
  * pheatmap
  * tidyverse
  * hrbrthemes
  * viridis
  * vcd


```{r packages, echo = FALSE, warning = FALSE, message = FALSE}

library(corrplot)
library(RColorBrewer)
library(pheatmap)
library(tidyverse)
library(hrbrthemes)
library(viridis)
library(vcd)

```

Set Working Directory

```{r wd, echo = FALSE, warning = FALSE, message = FALSE}

mywd <- "~/Library/CloudStorage/Dropbox/github/SPN4_maternal_mRNA/04_Tintori_et_al_Comparison_Fig_2_S2"
```


## Data Import and Munging

#### Import Metadata from Tintori et al data

This metadata describes each sample in Tintori et al., 

```{r import1, echo=TRUE, warning = FALSE, message = FALSE}

###########  READ IN THE DATA  #####################

# Set dir
setwd(mywd)

# Import the metadata from Tintori et al.
metadata_tintori <- read.table(file = "01_input/fromTintori/metadata_parsed_50306.txt", header = FALSE, fill = TRUE)

# Annotate & Resort the metadata
colnames(metadata_tintori) <- c("sample", "cellID", "stage", "rep")

# Re-order by sample name
metadata_tintori <- metadata_tintori[order(metadata_tintori$sample),]
dim(metadata_tintori)
head(metadata_tintori)
```

#### Import Tintori et al Normalized RNA-seq data

```{r import2, echo = TRUE, message = FALSE, warning = FALSE}

# Set dir
setwd(mywd)

# Import the normalized RNA-seq data from Tintori et al.
tintori <- read.table(file = "01_input/fromTintori/TableS2_RPKMs.txt", header = TRUE)
dim(tintori)

# Transform the RNA-seq data long-wise using tidyverse
tintori_long <- pivot_longer(tintori, cols = 2:220, names_to ="sample", values_to = "rpkm")
dim(tintori_long)
head(tintori_long)
```

#### Join the metadata and RNA-seq data


```{r join, echo = TRUE, warning = FALSE, message = FALSE}
# Join the Noramlized RNA-seq data to the metadata into tintori_merged
tintori_merged <- left_join(tintori_long, metadata_tintori)

# Remove tossed cellIDs
dim(tintori_merged)
head(tintori_merged)

# How many samples should be tossed?
length(which(metadata_tintori$cellID == "tossed"))
# 54

# How many entries in the merged set?
length(which(tintori_merged$cellID == "tossed"))
# 1,694682

# Remove "tossed" entries
tintori_retained_merged <- tintori_merged[which( !tintori_merged$cellID == "tossed"),]
dim(tintori_retained_merged)
head(tintori_retained_merged)

# Reduce the data structue down into a nested data frame:
tintori_nest_by_transcript <- tintori_retained_merged |>
  group_by(transcript) |>
  nest() 

```

#### Import Wormbase identifiers

```{r import3, echo = TRUE, message = FALSE, warning = FALSE}

# Merge with a list of WormBase IDs:

# Set dir
setwd(mywd)

# Import the list that I downloaded from simplemine. There is a screenshot in the same folder that shows what I selected
wormbaseIDs <- read.table(file = "01_input/fromSimpleMine/simplemine_results_WBGene_publicName.txt", header = TRUE, fill = TRUE)
head(wormbaseIDs)

```
#### Merge the tintori data with the Wormbase Gene identifiers

```{r merge, echo = TRUE, message = FALSE, warning = FALSE}

# Merge the annotation genenames to the tintori nested data frame:          
tintori_nest_by_transcript <- left_join(tintori_nest_by_transcript, wormbaseIDs, join_by( transcript == Public_Name ), keep = TRUE) %>%
  select( c("transcript", "WormBase_Gene_ID", "Public_Name", "data"))
  
dim(tintori_nest_by_transcript)
head(tintori_nest_by_transcript)

```

----

## Heatmaps 

```{r makematrix1, echo = FALSE, warning = FALSE, message = FALSE}

#### Make a giant heatmap ####

# Filter for present genes greater than 5 summed across 
tintori_nest_present <- tintori_nest_by_transcript |>
  rowwise() |>
  mutate(total_rpkm = sum(data$rpkm)) |>
  filter(total_rpkm > 5)

# I have 14,776 genes that passed the filter
dim(tintori_nest_present)
head(tintori_nest_present)


# Filter for changing genes that have a variance over 10 
changing <- tintori_nest_present |>
  rowwise() |>
  mutate(var = 100*var(data$rpkm)/total_rpkm) |>
  filter(var > 10)  

unchanging <- tintori_nest_present |>
  rowwise() |>
  mutate(var = 100*var(data$rpkm)/total_rpkm) |>
  filter(var <= 10)  

# I have 12,783 changing genes
dim(changing)

# I have 1,993 unchanging genes
dim(unchanging)

# Unnest the changing datastructure and convert into a matrix for heatmap plotting
# Unnest
changing_unnested <- changing |>
  select(transcript, data) |>
  unnest(data) |>
  select(transcript, cellID, rpkm, rep, stage) |>
  mutate( cell_combo = paste(cellID, rep, sep = "_"))
#changing_unnested

# Here is a dataframe that has each column as a cell type + replicate and each row as a gene and each value as rpkm
changing_unnested_wide <- changing_unnested |>
  filter(stage != "16-cell") |>
  select(transcript, rpkm, cell_combo) |>
  pivot_wider(names_from = cell_combo, values_from = rpkm)
#changing_unnested_wide

# Here is a dataframe that merges each column as a cell type (replicates merged), each row is a gene, and each value is the mean rpkm
changing_wide_df <- changing_unnested |>
  group_by(transcript, cellID) |>
  summarise(mean_rpkm = mean(rpkm)) |>
  pivot_wider(names_from = cellID, values_from = mean_rpkm)
#changing_wide_df
dim(changing_wide_df)
#str(changing_wide_df) #it's a tibble data frame

# change to a data frame
changing_wide_df <- as.data.frame(changing_wide_df)
#changing_wide_df

# create rownames
rownames(changing_wide_df) <- changing_wide_df$transcript

# Remove transcript column
changing_wide_df <- changing_wide_df[ 2:dim(changing_wide_df)[2]]
changing_wide_df <- as.data.frame(changing_wide_df)
#str(changing_wide_df)
#dim(changing_wide_df)

# Change to a matrix:
changing_wide_mat <- as.matrix(changing_wide_df)
```


Import the prefered cell order

```{r import4, echo = TRUE, message = FALSE, warning = FALSE}

# Set dir
setwd(mywd)

# Order the cell types by stage and anterior-> posterior like so:
cellType_sorted <- read.table(file = "01_input/cellTypes_sorted.txt", header = TRUE)
cellType_sorted

```



```{r makematrix2, echo = FALSE, message = FALSE, warning = FALSE}

colnames(changing_wide_mat)

bestOrder <- c(23, 1, 24, 2, 7, 17, 25, 3, 5, 8, 10, 12, 16, 20, 26, 4, 6, 9, 11, 13, 14, 15, 18, 19, 21, 22, 27)

# Do I have 27 cell types?
length(bestOrder)
length(unique(bestOrder))
dim(cellType_sorted)[1]

# re-arrange changing_wide_mat columns by the best order 
changing_wide_mat <- changing_wide_mat[ , bestOrder]

#changing_wide_mat

# changing_wide_mat is my matrix. Next steps will go on to make the heatmaps:

```

Not included: optimize heatmap type

Included: Cannaberra complete heatmap

```{r heatmap1, echo = TRUE, message = FALSE, warning = FALSE}


# Euclidean also looks good
# "ward.D", "ward.D2", "single", "complete", "average" (= UPGMA), "mcquitty" (= WPGMA), "median" (= WPGMC) or "centroid" (= UPGMC).
#wardD - looks good
#wardD2 - looks good
#single- nope

# OF all these, the canberra complete looks the best at separataing out the "decay" group. Euclidean Complete is also very good.

# Here is my favorite heatmap: cannaberra + complete. this one is great for separating out the zygotically transcribed (global) and maternal decay (global) genes:
set.seed(2025)
heatmap_caco5 <- pheatmap(changing_wide_mat, 
         cluster_cols = FALSE,
         scale = "row", 
         show_rownames = FALSE,
         clustering_distance_rows = "canberra", 
         clustering_method = "complete", 
         cutree_rows = 5)

# I split it into five clusters. I want to annotate the cluster number onto the plot so we can ensure that the cluster numbers match. 
# I want what looks like the 2nd and 3rd clusters as the "maternal transcripts undergoing decay groups"

# obtain the cluster information and add it into the matrix
cluster_set = cutree(heatmap_caco5$tree_row, 5)
clustered_changing_wide_mat <- cbind(changing_wide_mat, 
                     cluster = cluster_set )

# add cluster data into its own data frame
ann = data.frame(cluster_set)
rownames(ann) = rownames(clustered_changing_wide_mat)

#head(ann)
# Colors of the 
vcolors = plasma(7)[1:5]
#vcolors[1]


my_colour = list(cluster_set = c( "1" = vcolors[1], "2" = vcolors[5], "3" = vcolors[3], "4" = vcolors[2], "5" = vcolors[4]))


# Make an annotated heatmap using canaberra + complete:
set.seed(2025)
heatmap_caco5_ann <- pheatmap(changing_wide_mat, 
         cluster_cols = FALSE,
         scale = "row", 
         annotation_colors = my_colour,
         show_rownames = FALSE,
         clustering_distance_rows = "canberra", 
         clustering_method = "complete", 
         cutree_rows = 5,
         annotation_row=ann)

```

Save the heatmap to a file:

```{r heatmap2, echo = TRUE, message = FALSE, warning = FALSE}

# Set dir
setwd(mywd)

# Save this plot: 
today <- format(Sys.Date(), "%y%m%d")
filename1 <- paste("03_output/", today, "_Heatmap_Tintori_changing_present_transcripts_clust_ann.pdf", sep = "")
pdf(filename1, height = 8, width = 6)

heatmap_caco5_ann
dev.off()

# OK, so the gene clusters I want are clusters 5 and 3: 

```

---- 

## SAVE CLUSTER LISTS - PLOT LINEPLOTS BY CLUSTER LIST

```{r clusterlists, echo = FALSE, message = FALSE, warning = FALSE}

# filter out lists of genes from each cluster
head(clustered_changing_wide_mat)

# change to a data frame:
clustered_changing_wide_df <- as.data.frame(clustered_changing_wide_mat)
#head(clustered_changing_wide_df)

# Set the cluster # as a factor:
clustered_changing_wide_df$cluster <- as.factor(clustered_changing_wide_df$cluster)

# How many in each cluster:
summary(clustered_changing_wide_df$cluster)

# subset clusters
cluster1 <- clustered_changing_wide_df |>
  filter(cluster == 1)
cluster2 <- clustered_changing_wide_df |>
  filter(cluster == 2)
cluster3 <- clustered_changing_wide_df |>
  filter(cluster == 3)
cluster4 <- clustered_changing_wide_df |>
  filter(cluster == 4)
cluster5 <- clustered_changing_wide_df |>
  filter(cluster == 5)

# Create a dataframe that is grouped by the cluster and calculate the mean values of each cell ID across the cluster:
mean_by_cluster <- clustered_changing_wide_df %>% 
  group_by(cluster) %>%
  summarise_all("mean")

# eda 
dim(mean_by_cluster)
meanbycluster <- as.data.frame(mean_by_cluster)
meanbycluster

# convert the meanbycluster into longform data to plot using ggplot2:
longer_means_by_cluster <- pivot_longer(meanbycluster, cols = 2:28, names_to ="cell",
             values_to = "intensity")

# Set cell ID as an ordered factor
longer_means_by_cluster$cell <- factor(longer_means_by_cluster$cell, levels = colnames(changing_wide_mat))

longer_means_by_cluster$cluster <- factor(longer_means_by_cluster$cluster, levels = c(1,4,3,5,2))

# create lineplots of the trend over cell type for each cluster
vcolors = plasma(7)[1:5]
#vcolors
lineplot_clusters <- ggplot(longer_means_by_cluster, aes(x=cell, y=intensity, group=cluster, colour=cluster)) + 
  geom_line()+
  geom_point() +
  scale_color_manual(values = vcolors) +
  facet_grid(rows = vars(cluster))

lineplot_clusters
```

# Save theis plot to annotate the heatmap:

```{r heatmap_and_clusters, echo = TRUE, message = FALSE, warning = FALSE}

setwd(mywd)

today <- format(Sys.Date(), "%y%m%d")
filename2 <- paste("03_output/", today, "_cluster_lineplots.pdf", sep = "")
pdf(filename2, width = 6, height = 6)

lineplot_clusters

dev.off()

```



#* import the gene lists of SPN-4, OMA-1, and LIN-41 targets

```{r import5, echo = TRUE, warning = FALSE, message = FALSE}

setwd(mywd)
# import the gene lists
OMA1_1 <- read.table(file = "01_input/tlists_from_02/20250711_OMA1_ONLY_list.txt")
SPN4_2 <- read.table(file = "01_input/tlists_from_02/20250711_SPN4_ONLY_list.txt")
LIN41_3 <- read.table(file = "01_input/tlists_from_02/20250711_LIN41_ONLY_list.txt")
OMA1LIN41_4 <- read.table(file = "01_input/tlists_from_02/20250711_OMA1_LIN41_list.txt")
OMA1SPN4_5 <- read.table(file = "01_input/tlists_from_02/20250711_OMA1_SPN4_list.txt")
LIN41SPN4_6 <- read.table(file = "01_input/tlists_from_02/20250711_LIN41_and_SPN4_list.txt")
ALL_7 <- read.table(file = "01_input/tlists_from_02/20250711_OMA1_SPN4_LIN41_list.txt")

20250711_LIN41_and_SPN4_list.txt
20250711_LIN41_list.txt
20250711_LIN41_ONLY_list.txt
20250711_OMA1_LIN41_list.txt
20250711_OMA1_list.txt
20250711_OMA1_ONLY_list.txt
20250711_OMA1_SPN4_LIN41_list.txt
20250711_OMA1_SPN4_list.txt
20250711_SPN4_list.txt
20250711_SPN4_ONLY_list.txt
20250715_Presentlist.txt

# eda
str(OMA1_1)
str(SPN4_2)
str(LIN41_3)
str(OMA1LIN41_4)
str(OMA1SPN4_5)
str(LIN41SPN4_6)
str(ALL_7)

# annotate the dataframes:
OMA1_1 <- OMA1_1 %>% 
  mutate(IP = "OMA-1")

SPN4_2 <- SPN4_2 %>% 
  mutate(IP = "SPN-4")

LIN41_3 <- LIN41_3 %>% 
  mutate(IP = "LIN-41")

OMA1LIN41_4 <- OMA1LIN41_4 %>% 
  mutate(IP = "OMA-1_LIN-41")

OMA1SPN4_5 <- OMA1SPN4_5 %>% 
  mutate(IP = "OMA-1_SPN-4")

LIN41SPN4_6 <- LIN41SPN4_6 %>% 
  mutate(IP = "LIN-41_SPN-4")

ALL_7 <- ALL_7 %>% 
  mutate(IP = "OMA-1_SPN-4_LIN-41")
ALL_7

# merge the gene lists
IP_lookup <- rbind(OMA1_1, SPN4_2, LIN41_3, OMA1LIN41_4, OMA1SPN4_5, LIN41SPN4_6, ALL_7)
colnames(IP_lookup) <- c("WormBase_Gene_ID", "IP") 
dim(IP_lookup)
table(IP_lookup$IP)
head(IP_lookup)

# user wormbaseID dataframe to add transcript names to gene lists
head(wormbaseIDs)
IP_lookup_annotated <- left_join(IP_lookup, wormbaseIDs)
head(IP_lookup_annotated)
# merge gene lists with changing_wide_mat

# create an annotated heatmap
changing_wide_df <- rownames_to_column(as.data.frame(changing_wide_mat))
head(changing_wide_df)

changing_wide_IP <- left_join(changing_wide_df, IP_lookup_annotated, by = c("rowname"="Sequence_Name"))

changing_wide_df_plusIP <- column_to_rownames(changing_wide_IP[,1:30]) 
# Add the column changing_wide_IP$IP to the annotation table and re-map the heatmap 

# add cluster data into its own data frame
head(ann)
head(changing_wide_df_plusIP)
table(rownames(ann) == rownames(changing_wide_df_plusIP))

changing_wide_df_plusIP
ann2 <- cbind(ann, changing_wide_df_plusIP$IP)
colnames(ann2) <- c("cluster_set", "IP")

ann3 <- ann2