---
title: "250714_GOontology_SPN4_mRNAs"
author: "Erin Osborne Nishimura"
date: "2025-07-14"
output: github_document
---

## Goal of the project

I want to detemrine which GO Ontology categories are enriched in the set of SPN-4 associated mRNAs and other categories of mRNAs. We have SPN-4, LIN-41, and OMA-1 associated mRNA categories. What GO ontology categories are over-represented in these?

Note - what is the best background data? I tried two different “universe” or background datasets. 1) all present genes or 2) all Worm Base genes (default for Cluster Profiler). In the end, we chose the Worm Base geneset as a “universe” as this best addressed the question: what gene categories are enriched in the SPN-4, OMA-1, and LIN-41 associated mRNAs? 

## Load modules

Load the following modules: 

  - clusterProfiler
  - enrichplot
  - ggplot2


```{r echo=TRUE, warning = FALSE, message = FALSE}

# Optional installation for cluster profiler:
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("clusterProfiler")

# Load libraries:
library(clusterProfiler)
library(enrichplot)
library(ggplot2)

```

## Initiate Cluster Profiler

Select org.Ce.eg.db from [Bioconductor Release (3.21), Bioconductor Annotation Packages](https://bioconductor.org/packages/3.21/data/annotation/).

```{r load_elegans, echo=TRUE, warning = FALSE, message = FALSE}

# BiocManager::install("org.Ce.eg.db")
library(org.Ce.eg.db)
organism = "org.Ce.eg.db"

```

## Import datasets 

These datasets are lists that were generated in a previous analysis located in 02_SPN4_LIN41_OMA1_Comparison_Fig_1_S1/04_output_data located within this same github repository

```{r import, echo=TRUE, warning = FALSE, message = FALSE}

# set wd:
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")
getwd()

# import the data files
SPN4_list <- read.table(file = "01_input/20250711_SPN4_list.txt", quote = "")
SPN4_list2 <- sort(SPN4_list$V1, decreasing = TRUE)

OMA1_list <- read.table(file = "01_input/20250711_OMA1_list.txt", quote = "")
OMA1_list2 <- sort(OMA1_list$V1, decreasing = TRUE)

LIN41_list <- read.table(file = "01_input/20250711_LIN41_list.txt", quote = "")
LIN41_list2 <- sort(LIN41_list$V1, decreasing = TRUE)

Present_list <- read.table(file = "01_input/20250715_Presentlist.txt", quote = "")
Present_list2 <- sort(Present_list$V1, decreasing = TRUE)

# merge datasets into a list of vectors:
RBP_list_of_vectors <- list(SPN4 = SPN4_list2, OMA1 = OMA1_list2, LIN41 = LIN41_list2)
```

## Biological Processes 

### Compare the clusters

```{r compare, echo=TRUE}

# Compare the clusters 
compareRBPclusters_BP <- compareCluster(geneCluster = RBP_list_of_vectors, fun = 'enrichGO', OrgDb = 'org.Ce.eg.db', ont = "BP", keyType="WORMBASE", pAdjustMethod = "BH", qvalueCutoff = 0.02)

#universe =  Present_list2,
compareRBPclusters_BP <- setReadable(compareRBPclusters_BP, OrgDb = 'org.Ce.eg.db', keyType="WORMBASE")

## S4 method for signature 'enrichResult'
compareRBPclusters_BP_simple <- simplify( compareRBPclusters_BP,
          cutoff = 0.6,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang",
          semData = NULL)

compareRBPclusters_BP
compareRBPclusters_BP_simple

# Plot the clusters
dotplot(compareRBPclusters_BP)
dotplot(compareRBPclusters_BP_simple)
help(dotplot)


```

### Save the plot

Print Biological Processes plot to a file. Use this in Supplemental Figure. 

```{r printBP, echo=TRUE, message = FALSE, warning = FALSE}

# Save the plot 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

pdf("03_output/SPN4_OMA1_LIN41_dotPlot_BP.pdf", height = 10, width = 10)
cluster.p1 <- dotplot(compareRBPclusters_BP_simple, 
              title = "GO Enrichment (Biological Process)")
cluster.p1 + theme(
  axis.text.y = element_text(size = 15),
  axis.text.x = element_text(size = 15),
  legend.text = element_text(size = 15)
  )

dev.off()
```
## Identify key genes 

Determine the genes that are in each GO ontology category

Use these gene names in the Supplemental Figure

```{r BPdata, echo = FALSE, message = FALSE}

# Total lists of genes
compareRBPclusters_BP_simple@compareClusterResult$Description

# Lists of genes - set 1
compareRBPclusters_BP_simple@compareClusterResult$Description[1]
compareRBPclusters_BP_simple@compareClusterResult$geneID[1]
compareRBPclusters_BP_simple@compareClusterResult$Count[1]

# Lists of genes - set 2
compareRBPclusters_BP_simple@compareClusterResult$Description[2]
compareRBPclusters_BP_simple@compareClusterResult$geneID[2]
compareRBPclusters_BP_simple@compareClusterResult$Count[2]

# Lists of genes - set 3
compareRBPclusters_BP_simple@compareClusterResult$Description[3]
compareRBPclusters_BP_simple@compareClusterResult$geneID[3]
compareRBPclusters_BP_simple@compareClusterResult$Count[3]

# Lists of genes - set 4
compareRBPclusters_BP_simple@compareClusterResult$Description[4]
compareRBPclusters_BP_simple@compareClusterResult$geneID[4]
compareRBPclusters_BP_simple@compareClusterResult$Count[4]

# Lists of genes - set 5
compareRBPclusters_BP_simple@compareClusterResult$Description[5]
compareRBPclusters_BP_simple@compareClusterResult$geneID[5]
compareRBPclusters_BP_simple@compareClusterResult$Count[5]

# Lists of genes - set 17
compareRBPclusters_BP_simple@compareClusterResult$Description[12]
compareRBPclusters_BP_simple@compareClusterResult$geneID[12]
compareRBPclusters_BP_simple@compareClusterResult$Count[12]
```

## Save the datasets 

Save the datasets so that readers can have access to the sets of genes within each category. 

```{r Dataset, echo = TRUE}
# output the list 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

write.csv(compareRBPclusters_BP_simple@compareClusterResult, file = "03_output/SPN4_assocRNAs_GO_Ontology.csv",
          quote = FALSE)

```

## Molecular Function

### Compare the clusters

```{r compareMF, echo=FALSE, warning = FALSE, message = FALSE}

# Compare the clusters 
compareRBPclusters_MF <- compareCluster(geneCluster = RBP_list_of_vectors, fun = 'enrichGO', OrgDb = 'org.Ce.eg.db', ont = "MF", keyType="WORMBASE", pAdjustMethod = "BH", qvalueCutoff = 0.02)
#universe =  Present_list2, 
compareRBPclusters_MF <- setReadable(compareRBPclusters_MF, OrgDb = 'org.Ce.eg.db', keyType="WORMBASE")

## S4 method for signature 'enrichResult'
compareRBPclusters_MF_simple <- simplify(compareRBPclusters_MF,
          cutoff = 0.6,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang",
          semData = NULL)

#compareRBPclusters_MF
#compareRBPclusters_MF_simple

# Plot the clusters
dotplot(compareRBPclusters_MF)
dotplot(compareRBPclusters_MF_simple)
```

### Save the plot

Print the Molecular Function to a file

```{r printMF, echo=TRUE, message = FALSE, warning = FALSE}

# Save the plot 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

pdf("03_output/SPN4_OMA1_LIN41_dotPlot_MF.pdf", height = 5, width = 10)
cluster.p1 <- dotplot(compareRBPclusters_MF_simple,
              title = "GO Enrichment (Molecular Function)")
cluster.p1 + theme(
  axis.text.y = element_text(size = 15),
  axis.text.x = element_text(size = 15),
  legend.text = element_text(size = 15)
  )

dev.off()

```

## Cellular Components

### Compare the clusters


```{r compareCC, echo=FALSE, warning= FALSE, message = FALSE}

# Compare the clusters 

compareRBPclusters_CC <- compareCluster(geneCluster = RBP_list_of_vectors, fun = 'enrichGO', OrgDb = 'org.Ce.eg.db', ont = "CC", keyType="WORMBASE", pAdjustMethod = "BH", qvalueCutoff = 0.02)
# universe =  Present_list2
compareRBPclusters_CC <- setReadable(compareRBPclusters_CC, OrgDb = 'org.Ce.eg.db', keyType="WORMBASE")

## S4 method for signature 'enrichResult'
compareRBPclusters_CC_simple <- simplify( compareRBPclusters_CC,
          cutoff = 0.6,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang",
          semData = NULL)

#compareRBPclusters_CC
#compareRBPclusters_CC_simple

# Plot the clusters
dotplot(compareRBPclusters_CC_simple)
```

### Save the plot

Print Cellular Components to a file

```{r printCC, echo=TRUE, warning = FALSE, message = FALSE}

# Save the plot 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

pdf("03_output/SPN4_OMA1_LIN41_dotPlot_CC.pdf", height = 5, width = 10)
cluster.p1 <- dotplot(compareRBPclusters_CC_simple,
              title = "GO Enrichment (Cellular Component)")
cluster.p1 + theme(
  axis.text.y = element_text(size = 15),
  axis.text.x = element_text(size = 15),
  legend.text = element_text(size = 15)
  )

dev.off()

```

## ALL Categories

### Compare the clusters

All category ontologies

```{r compareALL, echo=FALSE, message = FALSE, warning = FALSE}

# Compare the clusters 

compareRBPclusters_ALL <- compareCluster(geneCluster = RBP_list_of_vectors, fun = 'enrichGO', OrgDb = 'org.Ce.eg.db', ont = "ALL", keyType="WORMBASE", pAdjustMethod = "BH", qvalueCutoff = 0.02)
# universe =  Present_list2
compareRBPclusters_ALL <- setReadable(compareRBPclusters_ALL, OrgDb = 'org.Ce.eg.db', keyType="WORMBASE")

## S4 method for signature 'enrichResult'
compareRBPclusters_ALL_simple <- simplify( compareRBPclusters_ALL,
          cutoff = 0.6,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang",
          semData = NULL)

#compareRBPclusters_ALL
#compareRBPclusters_ALL_simple

# Plot the clusters
dotplot(compareRBPclusters_ALL_simple)
```
### Save the plot

Print All categories plot to a file

```{r printALL, echo=TRUE, warning = FALSE, message = FALSE}

# Save the plot 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

pdf("03_output/SPN4_OMA1_LIN41_dotPlot_ALL.pdf", height = 5, width = 10)
cluster.p1 <- dotplot(compareRBPclusters_ALL_simple,
              title = "GO Enrichment (ALL Ontologies)")
cluster.p1 + theme(
  axis.text.y = element_text(size = 15),
  axis.text.x = element_text(size = 15),
  legend.text = element_text(size = 15)
  )

dev.off()
```
## SPN-4 on its own

I wanted to analyze the SPN-4 set of associated mRNAs in isolation away from the other RBPs. 

```{r spn4ONlY, echo = TRUE, warning = FALSE, message = FALSE}


ego_SPN4 <- enrichGO(gene = RBP_list_of_vectors$SPN4,
                universe = names(Present_list2),
                keyType ="WORMBASE", 
                OrgDb = org.Ce.eg.db,
                ont = "BP",
                pAdjustMethod = "BH",
                readable = TRUE)

ego_SPN4_simple <- simplify(ego_SPN4,
          cutoff = 0.55,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang",
          semData = NULL)

# Print
dotplot(ego_SPN4_simple)

```

## Save the plot

Print SPN-4 only plot to a file 

```{r printSPN4only, echo=FALSE, warning = FALSE, message = FALSE}

# Save the plot 
setwd("~/Dropbox/github/SPN4_maternal_mRNA/03_GO_Ontology_Fig_S1")

pdf("03_output/SPN4_only_BP.pdf", height = 5, width = 10)
cluster.p1 <- dotplot(ego_SPN4_simple,
              title = "GO Enrichment (Biological Processes)")
cluster.p1 + theme(
  axis.text.y = element_text(size = 15),
  axis.text.x = element_text(size = 15),
  legend.text = element_text(size = 15)
  )

dev.off()
```

## Session Info 


```{r sessionInfo, echo = TRUE}

sessionInfo()

```

